---
title: "Machine Learning"
output:
  html_document:
    df_print: paged
---

```{r}
evaluation <- function(prediction, target) {
    t <- table(prediction, target)[2:1,2:1]
    
    accuracy <- (t[1,1]+t[2,2])/(t[1,1]+t[1,2]+t[2,1]+t[2,2])*100
    
    recall <- (t[1,1]/(t[1,1]+t[2,1]))*100
    
    precision <- (t[1,1]/(t[1,1]+t[1,2]))*100

    cat("Accuracy:  ", accuracy, "% of all cases are correctly predicted by the model. \n")
    cat("Recall: ", recall, "% of the true 1s were also predicted by the model as 1s. \n")
    cat("Precision:   ", precision, "% of the predicted 1s are true 1s. \n")
    cat("Confusion matrix: ", "\n")
    t
}
```


```{r}
#download datasets, save them somewhere

setwd("C:/Users/grade/OneDrive/Studium/Berlin/Data Club/Fall 2019/8 Machine Learning")

load("income.RData")

```


```{r}
#explore the dataset
dim
head
```

```{r}
#randomly split the dataset
set.seed(12)
samples <- sample(nrow(income), 0.6*nrow(income))
samples
```

```{r}
#create a trainingset
trainingset <- income[samples, ]
trainingset
```

```{r}
#create a testset and look closer
testset <- income[-samples, ]
testset
```

```{r}
#separate the true values of the target variable in the test set
test_target <- income[-samples, 5]
test_target
```

```{r}
#build your model on the training set
training_model <- glm(Geschlecht ~ Haarlaenge, data = trainingset, family = "binomial")
```

```{r}
#test your model on the test set
test_probabilities <- predict(training_model, newdata = testset, type = "response")
test_probabilities
```


```{r}
#convert the probabilities into clear yes/no predictions
test_predictions <- ifelse(test= test_probabilities>0.5, yes = "W", no = "M")
test_predictions
```


```{r}
#evaluate your model
evaluation(test_predictions, test_target)
```



##Predicting fraud
```{r}
#load the data
transactions <- readRDS(file = "transactions.rds")
```

```{r}
# Check out the data
head(transactions)

# keep in mind that the dataset is quite big (2,770,409 observations!)
```

- isFraud - explains if transaction was a fraud or not (agent-based simulation), 1 is fraud,   0 is no fraud
- amount - amount of money in local currency
- oldbalanceOrg - account balance of sender before transaction
- newbalanceOrig - new account balance of sender after transaction
- oldbalanceDest - old account balance of receiver before transaction
- newbalanceDest - new account balance of receiver after transaction
- hour - states the hour of the day in which the transaction occurred
- type - type of transaction, 1 if transation was a CASH-OUT, 0 if it was a TRANSFER

```{r}
# First, split the dataset into a training and a test set
# Randomly save 80% of the observations in an object "samples"

set.seed(12) 
samples <- sample(nrow(transactions), 0.8*nrow(transactions))
```

```{r}
# Next, save all observations in "samples" into a separate training set

train <- transactions[samples,]

# Square brackets behind an object allow you to access observations (before the comma) and/or variables (behind the comma)
```

```{r}
# Now, save all transactions that are not in "samples" in a testset
# Advice: "-" excludes certain observations

test <- transactions[-samples,]
```

```{r}
# Next, you would like to save the true values of the target variable "isFraud" separately to be able to compare your predictions to them later on
# Advice: use the same technique as before but specify the variable as well

test_target <- transactions[-samples, 1]
```

```{r}
# Now it is time to build the model. For that, we take the logistic regression (the command is glm()). Save the model in an object. 
# Advice: if you want to use all variables to predict isFraud, you can use a "." behind the "~". / the argument family is "binomial" in our case. 

# Which dataset to we use to build the model?


training_model <- glm(isFraud~., data =  train, family = "binomial")

# This may take a minute
```

```{r}
# Now, we would like to test the trained model - which data do we use for that?
# Use the command predict() to do so


test_prob <- predict(training_model, 
                             newdata = test, 
                             type="response")

# Advice: specify the model and use "newdata" to specify which data you want to test on
```


```{r}
# Change the probabilities to clear yes/no predictions, use ifelse() to do so

test_pred <- ifelse(test = test_prob>0.5, yes = 1, no = 0)

# Advice: you need to specify three arguments: what you want to test, what the value shall become if the condition meets and what if it does not
```

```{r}
# Evaluate the model and use evaluation() for that

evaluation(test_pred, test_target)
```


### XGBoost works better in this case - demonstration
```{r}
library(caret)
library(caTools)
library(rpart)
library(xgboost)
```

```{r}
# Load the data
transactions_0 <- readRDS(file = "transactions1.rds")

# Splitting data into training and test sets
set.seed(1)
spl <- sample.split(transactions_0$isFraud, 0.8)
train_0 <- transactions_0[spl == TRUE,]
test_0 <- transactions_0[spl == FALSE,]
```

```{r}
# Converting the data into the form needed by XGBoost
# training set only x variables
x_train <- as.matrix(train_0[,-6])
# training set only y variable
y_train <- as.numeric(as.character(train_0$isFraud))

# test set only x variables
x_test <- as.matrix(test_0[,-6])
# test set only y variable - target variable to test on
test_target <- as.numeric(as.character(test_0$isFraud))

# the form XGBoost needs
train_XGB <- xgb.DMatrix(data = x_train, label = y_train)
test_XGB <- xgb.DMatrix(data = x_test, label = test_target)
```

```{r}
# Specify wanted parameters
parameter <- list(booster="gbtree", max_depth = 2, eval_metric= "error", objective = "binary:logistic")


# Build the model
set.seed(1)
training_model <- xgb.train(data= train_XGB, params=parameter, nrounds = 11, watchlist = list(test_set= test_XGB, train_set= train_XGB))
```

```{r}
# Application of model on test set
test_prob <- predict(training_model, newdata = test_XGB, type ="response")

#Converting probabilities into predictions
test_pred <- ifelse(test_prob > 0.5, 1, 0)
```

```{r}
evaluation(test_pred, test_target)
```

