---
title: "Programming Fundamentals"
author: "Marina Wyss"
date: "8/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Agenda
* Data types
    * Numeric/integers
    * Strings
    * Booleans
    * Factors
* Data structures
    * Vectors
    * Matrices
    * Data frames
    * Lists
* Comparison Operators
* Logical Operators
* if, else if, else statements
* for loops
* while loops
* functions
* Exercises
* Solutions

<i> Some examples in this worksheet are inspired by Pierian Data's Python and R courses on Udemy.</i><br>
<br>
This worksheet is available in both <b>R</b> and <b>Python</b> (R below of course), with some minor differences. We think it's important to become familiar with Python early on, since it is also a very popular language and will be necessary to know if you want to pursue a career with data. Some key points about the two:
<ul>
<li> R comes from stats, Python comes from computer science</li>
<li> So, R has many statistical methods that aren't available or are more challenging in Python, and Python has some machine learning/AI capabilities that aren't possible in R.</li>
<li> Python is a full-on programming language, with options far beyond data analysis.</li>
<li> R is more common in academia.</li>
</ul>
Neither is "harder" to learn than the other - it's more a matter of what you're used to. It makes sense to practice both as much as possible.

Boefore diving right into our agenda, we will quickly load a data set `hflights` from the `hflights` package. We will apply some of the knowledge learned from this session on the data set.
```{r}
#install.packages("hflights") #this line has been commented, you can remove the hash tag to uncomment and install the package
library(hflights)
data(hflights)
head(hflights, 10) #get first 10 rows of the data set
```

Integers: only whole numbers, though often R just considers them numerics. Often the distinction isn't really important.
### integer
```{r}
# include L at the end to tell R it's an integer and not normal numeric
fourInt <- 2L + 2L
class(fourInt)

#ascertain the data type of the year column of hflights
class(hflights$Year)
```


### numeric
Numerics: can be whole numbers or doubles/floats with decimals
```{r}
four <- 2 + 2
class(four)

flo <- 2.5 - 1.7
class(flo)
```

### string
Usually enclosed in single or double quotes.
Can be characters or numeric, but will be stored as characters.
```{r}
hi <- "Hello world"
class(hi)
typeof(hi) #you could also use typeof() to ascetain data type

numString <- "12345"
class(numString)

class(hflights$UniqueCarrier)
```

### booleans
True/false. Must be in caps. Can be either the whole word or just T/F, though the whole word is better form.
```{r}
TRUE
T

FALSE
F
```

### factors
Categorical variables.  They are useful when you want to display character vectors in a non-alphabetical order. 
```{r}
x <- c("Dec", "Mar", "Aug", "Sep")
typeof(x)
sort(x) #sorted alphabetically, but not the way you want

month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

y <- factor(x, levels = month)
sort(y) #now it sorts in 'monthly' sequential order

# can also create an ordered factor for ordinal variables
sizes <- c(rep("Large", 3), rep("Medium", 3), rep("Small",3))
sizes
sizes2 <- ordered(sizes)
str(sizes2)

sizes <- ordered(sizes, levels = c("Small", "Medium", "Large")) #change the order
str(sizes)
```

### vectors
One-dimensional vector that must have all the same element type. A sequence of data elements of the same data type.
```{r}
num <- 1:10
num

num * 2

vec <- rep("meow", 10)
```

### matrices
A matrix is a collection of data elements arranged in a two-dimensional rectangular layout. All columns in a matrix must have the same class (numeric, character, etc.) and the same length.
```{r}
mat <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
mat
 
mat_2 <- mat * 2 #numerical operation possible with matrices

# indexing a matrix; finding locationof an element in [row, column] order
mat[2, 1]
mat_2[1, 3]
```

### data frames
Basically a "spreadsheet". Two-dimensional. Key is that they can have different data types. Each column must be of one data type.
```{r}
names <- c('Tom', 'Nick', 'Sally', 'Joe')
ages <- c(20, 21, 19, 18)
gender <- c("Male", "Male", "Male", "Female")
myData <- data.frame(ages, names, gender)
myData
names(myData) <- c("Age", "Name", "Gender")

myData
```


Let's do some fun operations with our `hflights` dataset. We install the `dplyr` package and load the library.
```{r}
#install.packages("dplyr")

library(dplyr)

head(hflights) #check the first 6 rows

#check for the names of the columns in the data set
names(hflights)

#create a new data set, with three selected columns that you'd like to work with
hflights2 <- hflights %>% select(Year, Month, Dest) 
head(hflights2)
```

### lists
An ordered collection of objects (components). A list allows you to gather a variety of (possibly unrelated) objects under one name. For example, could have a few data frames, a matrix, etc.
```{r}
myList <- list(hflights2[1:5,], mat, x)
myList
```

#comparison operators
```{r}
1 > 2
1 < 2
1 >= 1
1 <= 4
1 == 1
'hi' == 'bye'
```
 
### logical operators

```{r}
# all statements must be true to return TRUE when using "and"
(1 > 2) & (2 < 3)

# only one statement must be true to return TRUE when using "or", denoted with |
(1 > 2) | (2 < 3)

(1 == 2) | (2 == 3) | (4 == 2)

names(hflights)

hflights3 <- hflights %>% select(Year, Month) %>% 
  filter(Month %in% c(1,2)) #get only the first and second months

head(hflights3)
tail(hflights3)

#check for the unique values of Month column to confirm that you have only months 1 and 2
unique(hflights3$Month) 
```

### if, else, if else
```{r}
if (1 < 2){ 
  print('yep!')
}

if (1 < 2){
    print('first')
} else {
    print('last')
}

if (1 == 2){
  print('first')
} else if(3 == 3) {
  print('middle')
} else {
  print('last')
}
```

### for loops
for each item, perform some action.
```{r}
for(hertie in hflights$Year[1:5]) { #print the first five rows of Year 
  print(hertie)
}

nums <- 1:5

for (item in nums){
  print(item)
}

for (item in nums){
  print("Yep")
}

# can use whatever word to iterate over
for (cake in nums){
  print(cake + cake)
}
```

### while loops
while loops are a way to have your program continuously run some block of code until a condition is met (or in other words, while a certain condition is true). 

A major concern when working with a while loop is to make sure that at some point the condition should become true, otherwise the while loop will go forever! 
```{r}
i <- 1

while (i < 5){
  print(paste0("i is: ", i))
  i <- i + 1
}
```

### functions
Formally, a function is a useful device that groups together a set of statements so they can be run more than once. They can also let us specify parameters that can serve as inputs to the functions.

On a more fundamental level, functions allow us to not have to repeatedly write the same code again and again. 
```{r}
# basic structure
functionName <- function(someInput) {
  # some action to perform
}

# example 1
square <- function(x){
  print(x^2)
}

square(2)

# example 2
greeting <- function(name){
    print(paste('Hi ', name))
}

greeting("dude")

# example 3
addNum <- function(num1, num2){
    print(num1 + num2)
}
```

Now, here's a more complicated example. It will return the lesser of two given numbers if both numbers are even, but returns the greater if one or both numbers are odd. <br>
For example: 
* lesser_of_two_evens(2, 4) would return 2
* lesser_of_two_evens(2, 5) would return 5

```{r}
lesser_of_two_evens <- function(a, b){
  if ((a %% 2) == 0 & (b %% 2) == 0){
    print(min(a, b))
  } else {
    print(max(a, b))
  }
}

lesser_of_two_evens(2, 4)
lesser_of_two_evens(2, 5)
```

## Exercises:
Now we're going to practice a bit! Here are some tasks:
<ol>
<li> Write a script that prints "Hello" if the variable X is equal to 1</li>
<li> Write a script that prints "Even Number" if the variable X is an even number, otherwise prints "Not Even"</li>
<li> White a script that prints the numbers 1 through 10 using a for loop. </li>
<li> White a script that prints the numbers 1 through 10 using a while loop. </li>
<li> Create a function that takes in a name as a string argument, and prints out "Hello name"</li>
<li> Create a function that will return the product of two integers.</li>
<li> Challenge one: Create a function that will return TRUE if an input integer is prime. Otherwise, return FALSE. </li>
</ol>

## Solutions

### 1
```{r}
x <- 1

if (x ==1){
    print("Hello")
}
```

### 2
```{r}
x <- 3 # change x to test

if (x %% 2 == 0){
    print('Even Number')
}else{
    print('Not Even')
}
```

### 3
```{r}
x <- 1

for (i in 1:10){
  print(x)
  x <- x + 1
}

```

### 4
```{r}
counter <- 1

while(counter < 11){
	print(counter)
	counter <- counter + 1
}

```

### 5
```{r}
helloYou <- function(name){
    print(paste('Hello',name))
}

helloYou("Jerry")
```

### 6
```{r}
prod <- function(num1, num2){
    return(num1 * num2)
}

prod(3, 4)
```

### 7
```{r}
primeCheck <- function(num){
  # Could put more checks for negative numbers etc...
  if (num == 2) {
    return(TRUE)
  }
  for (x in 2:(num - 1)){
      
    if ((num %% x) == 0){
      return(FALSE)
    }
  }
  return(TRUE)
}

primeCheck(2)
primeCheck(8)
primeCheck(11)
```


