---
title: "5 Exploratory Data Analysis"
output:
  html_document:
    df_print: paged
---


```{r}
library(tidyverse)
```

## Understanding the data

```{r}
library(titanic)
df <- titanic_train


#first overview
head(df)
#realization that we do not know what variables are -> understand the data
#look at it again and realize variables that seem mostly unique values, first look at size and structure


dim(df)
str(df) 
# realize character variables, also you may want to check unique values - e.g. by str() with factors (displays levels)
#View(df)
```

## See which data could be useful
```{r}
df$Name <- as.factor(df$Name)
# df$Name <- df$Name %>% as.factor()
str(df$Name)

df[, c("Name", "Sex", "Ticket", "Cabin", "Embarked")] <-  lapply(df[, c("Name", "Sex", "Ticket", "Cabin", "Embarked")], as.factor)
# lapply(df, mean)

str(df)
#realize: Name, Ticket have too many distinct values (too many to generalize upon)- yet Name could be used to identify families or see if title matters (feature engineering is not focus of this class)
# same goes for Cabin (yet could be used for approx position on ship)
df$PassengerId %>% unique() %>% length() #Same for PassengerId (no variation that helps explaining survival) - PassengerId and Ticket are even random
#realize: some NAs hiding in Embarked as "" 
```


```{r}
df <- df %>% select(-Name, -Ticket, -PassengerId)
str(df) #variables are out
```
## Check out the NAs
```{r}
#NAs
df[df$Embarked == "", "Embarked"] <- NA 
#NAs are important to look at because you might not have the data you think you have
colSums(is.na(df))

library(naniar)
vis_miss(df)

df <- df %>% select(-Cabin) #too many NAs, also too much to deal with now

# What to do with NAs here? 
# Generally you have two options: remove every row with a NA (you lose data) or impute the NAs with something that makes sense, e.g. its mean or a 0 (preferably s.th. better informed; you might bias your model)
# So be careful and curious! Especially in causal analysis: missingness might have a pattern (e.g. due to the data generating process, maybe useful); in prediction we care only about how we best inform our model 
# Models usually kick out NAs, so be aware of that

# We do not want to spend too much time on NAs so we just kick them out now (you shouldn't finalize that decision here)

# summary(df[is.na(df$Age)==TRUE,])

df <- df %>%  filter(is.na(df$Age)==FALSE)
df$Embarked %>% is.na() %>% sum()
df <- df %>% filter(is.na(df$Embarked)==FALSE)
dim(df)

```

```{r}
str(df)
```


## Each Variable by itself
```{r}
summary(df)
# Realize: 40% survived, more men than women, etc. but also the difference between factor and numerical variables

library(skimr)
skimr::skim(df)

plot1 <- ggplot(data = df, mapping = aes(x = df$Survived))
plot1 + geom_bar()


# facet example
df %>% gather() %>%
  ggplot(mapping = aes(x = value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(stat="count")

# df1 <- df %>% gather()
# Realize: distributions, possible outliers

```



## Check for Outliers
```{r}

a <- df %>% ggplot(mapping = aes(x=factor(0), y=Age)) + geom_boxplot()

b <- df %>% ggplot(aes(x=factor(0), y=Fare)) + geom_boxplot()
c <- df %>% ggplot(aes(x=factor(0), y=Parch)) + geom_boxplot()

library(ggpubr)
ggarrange(a,b,c)
```



## Relationship between the outcome variable and the others
```{r, message=FALSE}
library(GGally)
ggpairs(df)
df$Sex <- as.numeric(df$Sex)-1
ggpairs(df)

library(corrplot)
cplot <- df %>%
  select_if(is.numeric) %>%
  cor(use="complete.obs") %>%
  corrplot.mixed(tl.cex=0.85)

```
### Go into further analysis with Survival

#### First: Make a plot look beautiful (THEME)
```{r}
df$Survived <- as.factor(df$Survived)

#basis
p <- ggplot(data = df, mapping = aes(x=Pclass, fill=Survived)) + geom_bar(alpha=0.5)

#labels
p <- p +  ylab("Survival Rate")+ ggtitle("Survival Rate by Class")

#colours
library(wesanderson)
p <- p + scale_fill_manual(values = wes_palette("Darjeeling1"))

#theme (to adjust non-data components)
p + theme(legend.position = "top", axis.text=element_text(size=15), axis.text.x = element_text(angle = 90))

#given themes 
library(ggthemes)
p + theme_economist() 

p + theme_minimal()
```




```{r}
ggplot(data= df, mapping= aes(x= Fare, fill=Survived)) + geom_histogram(aes(y=..density..)) 
```


Here, the point is to get to know geom_point()
```{r}
p <- ggplot(data=df, mapping = aes(x= Age, y = Fare, color=Survived)) + geom_point()
p+ geom_smooth()
p + geom_smooth(method="lm", formula = y ~ poly(x, 3), se = FALSE)
```


## Do it yourself: relationship between Pclass and Survived and other questions you might have
```{r}
ggplot(data = df, mapping = aes(x=as.factor(Sex), fill=Survived)) + geom_bar()
```

Extended:
```{r}
library(scales)
ggplot(data = df, mapping = aes(x=as.factor(Sex), fill=Survived)) + geom_bar(aes(y= (..count..)/sum(..count..)))+scale_y_continuous(labels=scales::percent)
```

Extended: Alluvial Diagram
```{r}
library(alluvial)

tbl_summary <- df %>% 
  group_by(Survived, Sex, Pclass) %>%
  summarise(N = n()) %>% 
  ungroup

alluvial(tbl_summary[, c(1:3)],
         freq=tbl_summary$N, border=NA,
         col=ifelse(tbl_summary$Survived == "1", "blue", "gray"),
         cex=0.65,
         ordering = list(
         order(tbl_summary$Survived, tbl_summary$Pclass==1),
         order(tbl_summary$Sex, tbl_summary$Pclass==1),NULL))


```





